---
title: 'N741 Final Project: Urbanization and Suddent Infant Death'
author: "Taylor Thul"
date: '`r Sys.Date()`'
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knitr chunk options set to prevent
# code, warnings and messages from being
# shown in your final document
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

# libraries - remember to load all
# of the packages needed for your R code in this
# R markdown document (RMD) to run and compile
# here are a few to get you started
library(rmarkdown)
library(knitr)
library(N741pkg)
library(knitr)
library(car)
library(plyr)
library(ggplot2)
library(dplyr)
library(devtools)
library(DT)
library(xtable)
```

## Abstract

  Sudden Infant Death Syndrome (SIDS) kills thousands of infants year for reasons that remain unclear. Though there have been major advances in the prevention of SIDS, advancement has stalled in recent years.  To continue progress new risk factors must be identified to tailor prevention measures.  For this study we will investigate any associations between SIDS and urbanization. Utilizing data from the CDC failed to show evidence for a significant relationship between SIDS and urbanization.  We did, however, find an association for all types of infant death and urbanization.  This potential effect of city size and type on infant health is a promising area of future research.


## Introduction

_As a nurse, my background is in the neonatal intensive care unit (NICU).  In my capacity as a bedside nurse I often discharge patients to home, a process that includes extensive education regarding prevention of sudden infant death syndrome (SIDS).  SIDS, also known as crib or cot death, is the sudden, unexplained death of an infant younger than 1 year old after thorough investigation (CDC, 2017).  SIDS is the leading cause of death in children one month to one year of age.  It is also included as a cause of death in sudden unexpected infant death (SUID), a category that also includes accidental suffocation and strangulation in bed as well as those deaths where the cause remains unknown without a complete investigation (CDC, 2017). Of the 3,7000 SIUD in the United States in 2015, 43% were attributed to SIDS and 32% of unknown cause (CDC, 2017). SUID is a significant health concern and has been the focus of much research and major public health efforts.  Many risk factors have been identified including: prone sleeping, bed sharing, loose bedding, maternal smoking, and premature birth (Carlin & Moon, 2017).  A public education campaign, Safe to Sleep, launched in 1994 had great success in lowering the rate of SIDS death by more than 50% (NIH, 2017).  However since that time the rates have remained relatively unchanged and there are no clear answers regarding the etiology of the disease. 
_As a researcher, my interests focus on maternal/ infant health so I often encounter vital statistics and other data on SIDS reporting on a variety of variables.  However, I often wonder what other variables not reported may influence SIDS rates.  One is particular is if urbanization is at all associated with SIDS rates.  I would postulate more urban settings would facilitate greater exposure to education campaigns and other resources but also could increase exposure to air pollution and communicable diseases.  There could, of course, be many confounding variables in the proposed relationship but the first step is establishing if any association exists.
  Therefore the purpose of this paper is to explore any association between urbanization and sudden infant death syndrome (SIDS).  The null hypothesis is all levels of urbanization are equal (ie there is not relationship between urbanization and SIDS): H0: U1=U2=U3=U4=U5
  The alternative hypothesis is not all levels of urbanization are equal and there is a relationship between SIDS and urbanization.


## Dataset Description and Summary

_To explore the relationship between SIDS and urbanization I will utilize the CDC Multiple Cause of Death database.  It contains mortality for all US counties based on the death certificate and includes demographic data as well as ICD-10 codes for cause of death and urbanization level. Unfortunately the CDC Wonder website does not allow access to the entire dataset rather some variables must be selected and a subset of data will be generated for use.  To have access to the 2013 urbanization level I was not also able to have the 2006 urbanization levels or demographic variables such as ethnicity.  My final subset included: 

*2013 Urbanization Levels
  +Large Central Metro
  +Large Fringe Metro
  +Medium Metro
  +Small Metro
  +Micropolitan
  +NonCore (non-metro)
  
*Age Groups
  +< 1 day (Group 1)
  +1-6 days (Group 2)
  +7-27 days (Group 3)
  +28- 364 days (Group 4)
  
*ICD-10 Cause of Death
*Number of Deaths
*Population
*Crude Death Rate

_2013 urbanization levels were selected as they were the most current and the age category was chosen to be consistent with the definition of SIDS. The urbanization variable is categorical and includes large central metro, large fringe metro, medium metro, small metro, and micropolitan. The multilevel categorization is well suited for health related analyses because it differentiates counties within large metropolitan areas into “central” and “fringe” to capture the possible differences between a true urban environment compared to highly populous suburban areas.    The ICD-10 cause of death will give the most detailed way to extract death by SIDS as opposed to other causes.  I then did some data cleaning for the entire dataset including creating categorical variables for age group and urbanization level.  There was some missing data (labeled as "unavailable") in the crude death rates. Thankfully there was no missing data in overall number of deaths. I chose to omit the missing data from the analysis containing the crude deaths variables since the raw numbers would still be represented in number of deaths. The deaths and crude rate variable was greatly skewed to the right so I performed a log10 transformation on both of them.  Both were far more normally distributed after so the transformed variable was used for the remainder of analysis.  I created a subset of data isolating SIDS as a cause of death using the ICD-10 code R95.  This resulted in a small amount of observations (n=23) which seemed insufficient for the scope of this project.  Therefore I chose to modify my questions and look at the association between urbanization and overall infant deaths then more specifically death as a result of SIDS.  Additionally as many SUID deaths are unexplained including all causes of death may capture some of them as well, for example 'other and unspecified accidents.' 
_Overall the data was fairly clean and analysis could begin.  After some initial exploration it was noted there were some extreme outlines in the 'one day' age category.  Returning to the data I noted multiple large, outlining variables associated with prenatal and childbirth related deaths, such as 'certain conditions originating in the perinatal period.'  This cause of death is much more related to the prenatal period than infancy therefore I chose to eliminate the '1 day' age category from the overall analysis to avoid confounding with prenatal and birth related deaths.   


The CDC public databases located at: [https://wonder.cdc.gov/controller/datarequest/D77;jsessionid=4DC93A9C344C5A163A529948737DB795]( https://wonder.cdc.gov/controller/datarequest/D77;jsessionid=4DC93A9C344C5A163A5 29948737DB795)

## Dataset Summary and Visualization

#Summary of Overall Dataset 
```{r}
library(readxl)
#read in the data
COD <- read_excel("~/Box Sync/Multiple Cause of Death, 1999-2015.xls")
View(COD)

#just trying to see if anything I run will work
#to get collum names
names(COD)
summary(COD)

# there is some non numeric data in crude.rate, change to NA
COD[COD=="Unreliable"]<-""

# renaming variables
# df = dataframe
# old.var.name = The name you don't like anymore
# new.var.name = The name you want to get
names(COD)[names(COD) == 'Infant Age Groups Code'] <- 'age.group'
names(COD)[names(COD) == '2013 Urbanization Code'] <- 'urban.code'
names(COD)[names(COD) == 'Crude Rate'] <- 'crude.rate'
names(COD)[names(COD) == '2013 Urbanization'] <- 'urbanization'
names(COD)[names(COD) == 'UCD - ICD-10 130 Cause List (Infants) Code'] <- 'icd'

# recode deaths into categorical infant age groups
COD$age.cat <- revalue(COD$age.group, c("1d"="1", "1-6d"="2", "7-27d"="3", "28-364d"="4"))

# need to change crude rate to numeric variable
COD$crude.rate<-as.numeric(COD$crude.rate) 

```
#Summary of > 1 Day Subdata

```{r}
# using subset function to eliminate day 1 age group
COD2 <- subset(COD, age.cat > 1, 
select=c(urbanization:age.cat))
summary(COD2)
```

#Summary of SIDS Subdata

```{r}
#data looking at just SIDS
COD3 <- subset(COD, icd == "GR130-135", 
select=c(urbanization:age.cat))
summary(COD3)
```


The mean number of deaths is 

```{r}
mean(COD$Deaths)
```

The mean population is 
```{r}
mean(COD$Population)
```


### Summary Statistics Tables


```{r}
m <- COD
t1 <- N741pkg::tbl.continuous(m,m$crude.rate,"Crude Rate")
t2 <- N741pkg::tbl.continuous(m,m$Deaths,"Deaths")
t3 <- N741pkg::tbl.continuous(m,m$Population,"City Population")
knitr::kable(rbind(t1,t2,t3),
             caption = "Table of Summary Stats for Numeric Variables in Causes of Death Data")
```

```{r}
m <- COD3
t1 <- N741pkg::tbl.continuous(m,m$crude.rate,"Crude Rate")
t2 <- N741pkg::tbl.continuous(m,m$Deaths,"Deaths")
t3 <- N741pkg::tbl.continuous(m,m$Population,"City Population")
knitr::kable(rbind(t1,t2,t3),
             caption = "Table of Summary Stats for Numeric Variables in SIDS Deaths")
```


```{r}
m <- COD
gm <- dplyr::group_by(m,urbanization)
t1 <- N741pkg::tbl.cat(gm,gm$urbanization)
knitr::kable(t1,
             caption = "Frequency Table for Urbanization")

m <- COD
gm <- dplyr::group_by(m,`Infant Age Groups`)
t1 <- N741pkg::tbl.cat(gm,gm$`Infant Age Groups`)
knitr::kable(t1,
             caption = "Frequency Table for Infant Age Groups")

```
#For Frequency Table of ICD 10 Cause List of Death see Appendix 1

```{r}
m <- COD3
gm <- dplyr::group_by(m,urbanization)
t1 <- N741pkg::tbl.cat(gm,gm$urbanization)
knitr::kable(t1,
             caption = "Frequency Table for Urbanization of SIDS Data")

gm <- dplyr::group_by(m,`Infant Age Groups`)
t1 <- N741pkg::tbl.cat(gm,gm$`Infant Age Groups`)
knitr::kable(t1,
             caption = "Frequency Table for Infant Age Groups of SIDS Data")

```

#Aggregate Data Tables

```{r}
#aggregate data 
t2 <- aggregate(Deaths ~ urbanization + age.group, COD, mean)
knitr::kable(t2,
             caption = "Deaths By Urbanization and Age Group")

t3 <- aggregate(crude.rate ~ urbanization + age.group, COD, mean)
knitr::kable(t3,
             caption = "Crude Death Rate By Urbanization and Age Group")

#for SIDS
t4 <- aggregate(crude.rate ~ urbanization + age.group, COD3, mean)
knitr::kable(t4,
             caption = "SIDS Deaths By Urbanization and Age Group")

```


### Graphic Representation of Data 
*note all non-SIDS graphs are based on the data excluding infants < 1 day
```{r}
## using subset function 
COD2 <- subset(COD, age.cat > 1, 
select=c(urbanization:age.cat))

##transform deaths data
COD$deaths <- log10(COD$Deaths)
COD2$deaths <- log10(COD2$Deaths)
COD3$deaths <- log10(COD3$Deaths)
COD$tcrude.rate <- log10(COD$crude.rate)
COD2$tcrude.rate <- log10(COD2$crude.rate)
COD3$tcrude.rate <- log10(COD3$crude.rate)

#make the transformed rate a factor
COD$tcrude.rate<-as.numeric(COD$tcrude.rate)
COD2$tcrude.rate<-as.numeric(COD2$tcrude.rate)
COD3$tcrude.rate<-as.numeric(COD3$tcrude.rate) 

##histogram of new transformed deaths variable



##some exploratory analysis of the data
library(ggplot2)
ggplot(COD2, aes(x=Population, y=deaths, colour=as.factor(age.cat))) +
  geom_point() +
  geom_smooth(method=lm) +
  ggtitle("Deaths by Population Grouped by Age") +
  xlab("Population") +
  ylab("Deaths") +
  guides(colour = guide_legend(title = "Age Group"))

#new plot using data with age > 1 day  
ggplot(COD2, aes(x= age.cat, y=deaths, fill=urbanization)) +
  geom_boxplot() + 
  ggtitle("Deaths by Age Group") + 
  xlab ("Age Grouping") +
  ylab ("Log10 Deaths")

# Density plots with semi-transparent fill
ggplot(COD2, aes(x=deaths, fill=urbanization)) +
  geom_density(alpha=.3) +
  ggtitle("Density Plot of Death by Urbanization Level") +
  xlab ("Log10 Deaths") +
  ylab ("Density")

##COME BACK TO THIS IF TIME
#histogram of crude rate by urbanization
ggplot(COD2, aes(x=deaths, fill=urbanization)) +
  geom_histogram(position="identity") +
  ggtitle ("Deaths By Urbanization")
  xlab ("Log10 Deaths")
  
#don't use
#histogram of crude rate by age
#ggplot(COD2, aes(x=tcrude.rate, fill=urbanization)) +
  #geom_histogram(position="identity") +
  #ggtitle ("Rate of Deaths By Urbanization")
  #xlab ("Log10 Crude Rates")
  
#don't use
#scatterplots to look at data against one another
#ggplot(COD, aes(x=tcrude.rate, y=Population)) +
    #geom_point(shape=1)  

#plot some of the data 
library(ggplot2)
ggplot(COD2, aes(x = age.group, y = deaths)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) +
  ggtitle("Deaths by Age Group with Outliers") +
  xlab ("Age Group") +
  ylab ("Log10 Deaths")
  
##with SIDS data
ggplot(COD3, aes(x=Population, y=deaths, colour=as.factor(age.cat))) +
  geom_point() +
  geom_smooth(method=lm) +
  ggtitle("SIDS Rates by Population Grouped by Age") +
  xlab("Population") +
  ylab("Number of SIDS Deaths") +
  guides(colour = guide_legend(title = "Age Group"))

#explore SIDS data
#ggplot(COD3, aes(x=tcrude.rate, fill=urbanization)) + geom_density(alpha=.3)

ggplot(COD3, aes(x=deaths, fill=urbanization)) +
  geom_density(alpha=.3) +
  ggtitle("Density Plot of SIDS Data") +
  xlab ("Log10 Deaths")
  
#ggplot(COD3, aes(x=age.cat, y=deaths, fill=urbanization)) + geom_boxplot()


```

## Methods and Analysis 

Due to the smaller amount of data present when restricted to the subset of the SIDS ICD-10 code I chose to run analysis on both the overall data set and the subset.  Therefore my question then became if there was any association between urbanization level and infant death and more specifically death as a result of SIDS.  Since this was a continuous y and categorical x I chose to use ANOVA as the primary analysis method. I assumed the transformed data would behave as normal. 

#Output of Model for SIDS Subset of Data 
```{r}

#ANOVA for SIDS with age
AnovaModel.3 <- lm(crude.rate ~ age.cat*urbanization, data=COD, 
  contrasts=list(age.cat ="contr.Sum", urbanization ="contr.Sum"))
Anova(AnovaModel.3)
with(COD, (tapply(crude.rate, list(age.cat, urbanization), mean, 
  na.rm=TRUE))) # means
with(COD, (tapply(crude.rate, list(age.cat, urbanization), sd, 
  na.rm=TRUE))) # std. deviations
xtabs(~ age.cat + urbanization, data=COD) # counts
```

With the inclusion of age, urbanization is not a statistically significant variable in SIDS death. Age is significant, which makes sense because the age categories are grouped in a very uneven manner. Therefore I will reject the alternative hypothesis and accept the null.  No post-hoc testing is needed as the alternative is rejected. Next I will use ANOVA to analyze the full dataset (without day 1 death data).

#Output of Full Data Analysis
```{r}
#using COD2
AnovaModel.4 <- lm(Deaths ~ age.group*urbanization, data=COD2, 
  contrasts=list(age.group ="contr.Sum", urbanization ="contr.Sum"))
COD2Model <- Anova(AnovaModel.4)
with(COD2, (tapply(Deaths, list(age.group, urbanization), mean, 
  na.rm=TRUE))) # means
with(COD2, (tapply(Deaths, list(age.group, urbanization), sd, 
  na.rm=TRUE))) # std. deviations
xtabs(~ age.group + urbanization, data=COD2) # counts
```

```{r}
summary(AnovaModel.4)
summary(COD2Model)
plot(AnovaModel.4)
```
In this analysis both age and urban level are significant variables. However, given the large and varied causes of death and age groupings are.  This would be an interesting direction of future research but would require a separate analysis with the data arranged more appropriately for the question. 

```{r}
##logistic regression- I did this first but decided it was not entierly appropriate for the model
# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 
#set.seed(20170328)
#COD3$group <- runif(length(COD3$Deaths), min = 0, max = 1)


#COD3.train <- subset(COD3, group <= 0.80)
#COD3.test <- subset(COD3, group > 0.80)

#see if it worked
#summary(COD3.train)
#summary(COD3.test) 


#recoding categorical variables as factors
#COD3.train$urbanization.f <- factor(COD3.train$urbanization)
#is.factor(COD3.train$urbanization.f)

#is urbanization a predictor varaible?
#fit <- lm(Deaths ~ urbanization.f, data= COD3.train)
#summary(fit)

# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 

#set.seed(20170328)
#COD3$group <- runif(length(COD3$Deaths), min = 0, max = 1)

#what random forests do is this process over and over again and makes the aggregate which might be called bootstrapping?

#COD3.train <- subset(COD3, group <= 0.80)
#COD3.test <- subset(COD3, group > 0.80)

#see if it worked
#summary(COD3.train)
#summary(COD3.test) 


#recoding categorical variables as factors
#COD3.train$urbanization.f <- factor(COD3.train$urbanization)
#is.factor(COD3.train$urbanization.f)

#is urbanization a predictor varaible?
#fit <- lm(Deaths ~ urbanization.f, data= COD3.train)
#summary(fit)

```

```{r}
# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 

#set.seed(20170328)
#COD2$group <- runif(length(COD2$Deaths), min = 0, max = 1)

#creating the train and test dataset

#COD2.train <- subset(COD2, group <= 0.80)
#COD2.test <- subset(COD2, group > 0.80)

#see if it worked
#summary(COD2.train)
#summary(COD2.test) 

#are my categorical variables factors?
#is.factor(COD2$urbanization) #no this was not

#recoding categorical variables as factors
#COD2.train$urbanization.f <- factor(COD2.train$urbanization)
#is.factor(COD2.train$urbanization.f)

#is urbanization a predictor varaible?
#fit3 <- lm(Deaths ~ urbanization.f, data= COD2.train)
#summary(fit3)
```

## Discussion and Conclusions

After analysis there is not enough evidence to suggest an association between SIDS and urbanization levels.  However there is evidence that shows an association between all types of infant deaths and urbanization levels.  As this was not the original aim of the project it could be considered an incidental finding.  Perhaps the same factors hypothesizes as contributing to the theorized relationship between SIDS and urban level could have the same effect on overall infant deaths.  More urban areas could have higher risk of communicable diseases and more exposure to air pollutants, perhaps increasing risk for respiratory associated deaths.  However urban areas also often provide more accessibility to healthcare and public health education campaigns.  I would be interested in exploring this as a primary question in further analysis and investigating the direction of the relationship.  

## Limitations

There were many limitations to this analysis.  Access to the dataset was limited from the beginning, forcing exclusion of possibly significant variables such as ethnicity and presence of risk factors.  Furthermore once the SIDS specific data was manipulated it was far too small to do a quality analysis.  Use of a more varied and larger dataset would have vastly increased the validity of the findings.  

## References

Carlin RF, Moon RY. Risk Factors, Protective Factors, and Current Recommendations to Reduce Sudden Infant Death Syndrome A Review. JAMA Pediatric. 2017;171(2):175-180. doi:10.1001/jamapediatrics.2016.3345 

Centers for Disease Control and Prevention (2017 February 1). Sudden Unexpected Infant Death and Sudden Infant Death Syndrome Data and Statistics.  Retrieved from https://www.cdc.gov/sids/data.htm

Henderson and Velleman (1981), Building multiple regression models interactively. Biometrics, 37, 391–411.

##Github
The repository for this work may be found at (https://github.com/taylorthul/N741-Urban-SIDS-)[https://github.com/taylorthul/N741-Urban-SIDS-]

##Appendix
#Appendix 1
```{r}
m <- COD
gm <- dplyr::group_by(m,`UCD - ICD-10 130 Cause List (Infants)`)
t1 <- N741pkg::tbl.cat(gm,gm$`UCD - ICD-10 130 Cause List (Infants)`)
knitr::kable(t1,
             caption = "Frequency ICD- 10 Causes of Death")
```

