---
title: "SIDS Data Milestone 2 Draft"
date: "March 29, 2017"
output:
  word_document: default
  html_document: default
---
I am planning to use the template but I just wanted to try on my own at first. 

```{r setup, include=FALSE}
#get all the libraries and graphic options set
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(RcmdrMisc)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
library(knitr)
library(car)
library(plyr)
library(ggplot2)
library(dplyr)
library(devtools)
library(DT)
library(xtable)

```

Below is the section of R code to clean and arrrainge the data in a more useful way.  
```{r}
library(readxl)
#read in the data
COD <- read_excel("~/Box Sync/Multiple Cause of Death, 1999-2015.xls")
View(COD)

#just trying to see if anything I run will work
#to get collum names
names(COD)
summary(COD)

# renaming variables
# df = dataframe
# old.var.name = The name you don't like anymore
# new.var.name = The name you want to get
names(COD)[names(COD) == 'Infant Age Groups Code'] <- 'age.group'
names(COD)[names(COD) == '2013 Urbanization Code'] <- 'urban.code'
names(COD)[names(COD) == 'Crude Rate'] <- 'crude.rate'
names(COD)[names(COD) == '2013 Urbanization'] <- 'urbanization'
names(COD)[names(COD) == 'UCD - ICD-10 130 Cause List (Infants) Code'] <- 'icd'

# recode deaths into categorical infant age groups
COD$age.cat <- revalue(COD$age.group, c("1d"="1", "1-6d"="2", "7-27d"="3", "28-364d"="4"))

# there is some non numeric data in crude.rate, change to NA
COD[COD=="Unreliable"]<-""
COD

# need to change crude rate to numeric variable
COD$crude.rate<-as.numeric(COD$crude.rate) 
summary(COD)

#summary statistics
summary(COD)
mean(COD$Deaths)
mean(COD$Population)

```
Discussion explaining the summary statistics so I won't be useing the raw code. 

Now I will try to make some pretty tables:

```{r}
#this is where some code will go for tables that I have not quite yet figured out how to make becuase none of the code has worked so far 
```


These are some not-so-pretty tables that I am using just to look at the data and how it is related to one another. 
```{r}
#aggregate data (and not pretty tables)
aggregate(Deaths ~ age.group, COD, mean)
aggregate(Deaths ~ urban.code, COD, mean)
aggregate(crude.rate ~ age.group, COD, mean)
aggregate(crude.rate ~ urban.code, COD, mean)
aggregate(urban.code ~ Population, COD, mean)


```

Do some exploratory analysis and graphing (all of this will have explinations in the final draft).  I will not use all these graphs in the final report but I wanted to practice making them.  I will also go back in and tidy it up with lables and titles for each one chosen.  
```{r}

#some exploratory analysis of the data
#histogram
with(COD, Hist(crude.rate, scale="frequency", breaks="Sturges", col="red"))
#add lables
title (main= "Crude Rate of Deaths Overall")

#scatterplot
plot(COD$Population, COD$Deaths)

#histogram
with(COD, Hist(Deaths, scale="frequency", breaks="Sturges", col="red"))
#add lables
title (main= "Crude Number of Deaths Overall")

# histogram of deaths by urbanization
ggplot(COD, aes(x=Deaths, fill=urbanization)) +
    geom_histogram(binwidth=50, position="identity")

#histogram of crude rate by urbanization
ggplot(COD, aes(x=crude.rate, fill=urbanization)) +
    geom_histogram(position="identity")

#histogram of crude rate by age
ggplot(COD, aes(x=crude.rate, fill=age.cat)) +
    geom_histogram(position="identity")

#scatterplots to look at data against one another
ggplot(COD, aes(x=crude.rate, y=Population)) +
    geom_point(shape=1)  

#plot some of the data 
library(ggplot2)
ggplot(COD, aes(x = age.group, y = Deaths)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)

```

There are a lot of outliers in the one day age range.  This definately isn't SIDS. Going back into the data I see this is observation in row 28, which is deaths associated with 'Certain conditions originating in the perinatal period' 

```{r}
library(ggplot2)
ggplot(COD, aes(x = age.cat, y = crude.rate)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)
```

```{r}
#same plot but adding urbanziation varible 
ggplot(COD, aes(x=age.cat, y=age.cat, fill=urbanization)) + geom_boxplot()
```

I decide I shoud look at just the day 2- 365. This is really more of the informaiton I am interested in anyways and will help remove outliers.  I make a subset removing the first day deaths and explore it. 

```{r}
# using subset function 
COD2 <- subset(COD, age.cat > 1, 
select=c(urbanization:age.cat))
summary(COD2)

#new plot using data with age > 1 day  
ggplot(COD2, aes(x=age.cat, y=crude.rate, fill=urbanization)) + geom_boxplot()

# Density plots with semi-transparent fill
ggplot(COD2, aes(x=crude.rate, fill=urbanization)) + geom_density(alpha=.3)

#scatterplot
ggplot(COD2, aes(x=Deaths, y=urbanization, color=age.cat)) +
    geom_point(shape=1) +
    scale_colour_hue(l=50) + # Use a slightly darker palette than normal
    geom_smooth(method=lm,   # Add linear regression lines
                se=FALSE)    # Don't add shaded confidence region

```

This is intersting but I also want to just look at the SIDS data within the overall data set.
The ICD 10 code for SIDS is GR130-135 so I will use it to make a subset of the data looking at only SIDS across all time periods. 

```{r}
#data looking at just SIDS
COD3 <- subset(COD, icd == "GR130-135", 
select=c(urbanization:age.cat))
summary(COD3)

aggregate(crude.rate ~ age.group, COD3, mean)
aggregate(crude.rate ~ urban.code, COD3, mean)
aggregate(Deaths ~ urban.code, COD3, mean)

```

```{r}
#explore data
ggplot(COD3, aes(x=crude.rate, fill=urbanization)) + geom_density(alpha=.3)
ggplot(COD3, aes(x=Deaths, fill=urbanization)) + geom_density(alpha=.3)
ggplot(COD3, aes(x=age.cat, y=Deaths, fill=urbanization)) + geom_boxplot()

```

So I've done most of the exploratory analysis and cleaning at this point.  I will start making and testing models.  


```{r}
# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 

set.seed(20170328)
COD3$group <- runif(length(COD3$Deaths), min = 0, max = 1)

#what random forests do is this process over and over again and makes the aggregate which might be called bootstrapping?

COD3.train <- subset(COD3, group <= 0.80)
COD3.test <- subset(COD3, group > 0.80)

#see if it worked
summary(COD3.train)
summary(COD3.test) 


#recoding categorical variables as factors
COD3.train$urbanization.f <- factor(COD3.train$urbanization)
is.factor(COD3.train$urbanization.f)

#is urbanization a predictor varaible?
fit <- lm(Deaths ~ urbanization.f, data= COD3.train)
summary(fit)
```
Well none of this is significant. There were MANY fewer observation in the SIDS subset than in the whole thing so that may be contributing and I will adress this in my limitation section.  I may go back and re-code some of the other 'unknown' deaths that are considered SIDS but not classified as so to see if that gives more data but I have not decided quite yet. 

```{r}
#now looking at age category
is.factor(COD3$age.cat)

#recoding categorical variables as factors
COD3.train$age.f <- factor(COD3.train$age.cat)
is.factor(COD3.train$age.f)

#fit a model with all variables
fit2 <- lm(Deaths ~ age.f, data=COD3.train)
summary(fit2)# show results

```

Age category is a significant variable.  This, of course, makes tons of sense because age category 4 is much longer time (almost the whole year).  Does anything change when we put them together?

```{r}
#fit a model with all variables
fit3 <- lm(Deaths ~ age.f + urbanization.f, data=COD3.train)
summary(fit3)# show results
```

Well my theory is disproven. There does not seem to be any association between SIDS and urbanization.
Let me try modeling with the whole data set just because I have it already.  Does the association change when we look at all causes of death for infants between 2-364 days of age? 

```{r}
# divide the dataset into a training and a testing set based on a random uniform number on fixed seed, which in this case we are using the date
# this step is also creating a new variable and adding it to the data set which is a distribution of random numbers from 0 to 1 

set.seed(20170328)
COD2$group <- runif(length(COD2$Deaths), min = 0, max = 1)

#creating the train and test dataset

COD2.train <- subset(COD2, group <= 0.80)
COD2.test <- subset(COD2, group > 0.80)

#see if it worked
summary(COD2.train)
summary(COD2.test) 

#are my categorical variables factors?
is.factor(COD2$urbanization) #no this was not

#recoding categorical variables as factors
COD2.train$urbanization.f <- factor(COD2.train$urbanization)
is.factor(COD2.train$urbanization.f)

#is urbanization a predictor varaible?
fit3 <- lm(Deaths ~ urbanization.f, data= COD2.train)
summary(fit3)
```

Great so there is some significance here. I will explore/ explain this more in the final report.

```{r}

#recoding categorical variables as factors
COD2.train$age.f <- factor(COD2.train$age.cat)
is.factor(COD2.train$age.f)

#fit a model with all variables
fit4 <- lm(Deaths ~ age.f, data=COD2.train)
summary(fit4)# show results
```
Look at everything. 
```{r}
#fit a model with all variables
fit5 <- lm(Deaths ~ age.f + urbanization.f, data=COD2.train)
summary(fit5)# show results
```

Discussion. 
This is where the discussion will go! 


The git hub for this is located at (https://github.com/taylorthul/N741-Urban-SIDS-/blob/master/SIDS%20Data.Rmd) [https://github.com/taylorthul/N741-Urban-SIDS-/blob/master/SIDS%20Data.Rmd]